Parametricity Consequences of GADTs Representations

### 2021-02-22 20-gadts/GADT-story.tex
work on gadt-story

\section{Representations of Data Types}\label{sec:reps}

\subsection{Algebraic Data Types}\label{sec:adts}

@@

\subsubsection{Church Encodings of ADTs}

@@

\subsubsection{ADTs as Primitive Data Types}

@@

Primitive representations capture in syntax the fact that ADTs can be
considered as fixpoints. For example $\mathsf{List\,a}$ does indeed
satisfy the fixpoint equation $\mathsf{F\,X} = \mathsf{1 + a \times
  X}$ determined by~\eqref{eq:list}, i.e., is indeed such that
\[\mathsf{List\,a} = \mathsf{1 + a \times List\,a}\] 
since every element of $\mathsf{List\,a}$ is either empty or is
obtained by $\mathsf{Cons}$ing an element of type $\mathsf{a}$ onto an
already-existing structure of type $\mathsf{List\,a}$. In fact, this
fixpoint equation is just a rewriting of the Haskell data type
declaration for $\mathsf{List\,a}$. Similarly, the ADT
$\mathsf{Tree\,a}$ can be seen to be a solution to the fixpoint
equation
\[\mathsf{F\,X} = \mathsf{a + X \times a \times X}\] 
determined by~\eqref{eq:tree}.
%its Haskell data type declaration.

These fixpoint equations are entirely sensible at the level of
types. But to ensure that the syntactic fixpoint representing an ADT
actually denotes a semantic object computed as a real-live fixpoint,
the semantic fixpoint calculation must converge. This means that the
fixpoint being taken must be of {\em functors} on the category
interpreting types~\cite{tfca}, rather than of mere functions between
the sets interpreting types. Said differently, the function $F$
interpreting the type constructor $\mathsf{F}$ constructing the body
of a syntactic fixpoint must actually be a {\em functor} on the
category $\mathit{Set}$ of sets and functions. That is $F$ must not
only have an action on sets interpreting different type arguments to
$\mathsf{F}$, but must also have a {\em functorial action} on
functions between such sets. Reflecting this requirement back into
syntax gives that $\mathsf{F}$ must support a $\mathsf{map}$ function,
i.e., must be an instance of Haskell's $\mathsf{Functor}$ class.

Requiring $F$ to be a functor will ensure that the interpretation $\mu
F$ of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure
that $\mu F$ is itself a functor, so that the type constructor
$\mathsf{T}$ associated with the ADT $\mathsf{T\,a}$ supports a
$\mathsf{map}$ function, $F$ must itself be the fixpoint $\mu H$ of a
functor $H$. But the functor $H$ must be a functor on the category of
functors on the category $\mathit{Set}$ interpreting types. In other
words, $H$ must be a {\em higher-order} functor on
$\mathit{Set}$. Reflecting this requirement back into syntax gives
that $\mathsf{T\,a} = \mathsf{(\mu F)\,a} = \mathsf{(\mu\,(\mu
  H))\,a}$ for the ``type constructor constructor'' $\mathsf{H}$ whose
interpretation is $H$.

{\color{red} HERE!!!!}

A concrete example is given by the ADT $\mathsf{List\,a}$. This type
is is modelled as the fixpoint $\mu F_a$ of the first-order functor
whose action on objects is given by $F_a\,X = 1 + a \times X$ and,
whose action on morphisms (which models its corresponding
$\mathsf{map}$ function) is given by $F_a\,f = 1 + \id_a \times
f$. The type constructor $\mathsf{List}$ is then modelled by the
higher-order functor $H$ whose action on objects is given by $H\,F\,X
= 1 + X \times F\,X$ and whose action on morphisms is given by
$H\,F\,f = \id_1 + f \times F\,f$. (A {\em higher-order functor} in
this context is a functor not directly on the category interpreting
types, but rather on the category of functors interpreting type
constructors, i.e., on the category of functors on the category of
functors interpreting types.) The ADT $\mathsf{List\,a}$ is then
modelled as the functor $(\mu H)a$. Since $H$ is a higher-order
functor, not only does it have a corresponding functorial action,
i.e., support a function $\mathsf{map_H}$, but $\mu H$ is also a
functor with its own functorial action. That is, the type constructor
$\mathsf{List}$ itself supports a function $\mathsf{fmap_{List}}$.

{\color{blue} At the level of objects/types, short cut fusion can be
  used to show that the Church encoding and the primitive
  representation of an ADT are isomorphic.} Note, however, that
whereas it is impossible even to {\em state} the rearrange-transform
property for ADTs whose Church encodings do not have corresponding
$\mathsf{map}$ functions, and although this property is guaranteed to
hold even when it can be stated for Church encodings, it is simply the
reflection back into syntax of the naturality property of the
functorial action of the interpretation of the type constructor
associated with an ADT's representation as a primitive data type.

{\color{violet} Mention naturality? Semantics? Uniformity because we
  actually view ADTs as fixpoints of higher-order functors --- not
  first-order ones --- to get maps.}

%In the former case the ADT $\mathsf{List\,a}$ for each $\mathsf{a}$ is
%modelled as the fixpoint of its underlying functor $\mu F_a$; by
%contrast, in the latter case it is the type constructor
%$\mathsf{List}$ itself that is modelled as $\mu H$, and the ADT
%$\mathsf{List\,a}$ is actually modelled as $(\mu H)\,a$. That is, the
%ADTs $\mathsf{List\,a}$ for the various $\mathsf{a}$s are all modelled
%uniformly. (The extreme uniformity comes from the fact that the
%{\color{blue} type/second} argument $X$ is the same in each recursive
%call to $H$. This requirement will be lifted in
%Section~\ref{sec:gadts} below.)  This is precisely what makes it
%possible to define am $\mathsf{fmap}$ for list types.

%A crucial point is that while the $\mu F_a$ is not required to be a
%functor, and thus is not required to have a {\em functorial action} on
%morphisms --- i.e., to support an $\mathsf{fmap_{F_a}}$ function in
%programming terms --- the functor $\mu H$ modelling $\mathsf{List}$
%must do so. Indeed, since $H$ is a higher-order functor, $\mu H$ is
%itself a functor. Applying the functorial action $\mathsf{fmap_{\mu
%    H}}$ of this functor to a morphism $\mathsf{f}$ yields a natural
%transformation $\mathsf{fmap_{\mu H}\,f}$. This in turn guarantees the
%order-independence of type-independent rearrangement and conversion
%via $\mathsf{fmap_{\mu H}\,f}$ for lists {\color{blue} and trees}.

%{\color{blue} Extension of a container}

\begin{comment}
{\color{blue} Type-independent reasoning about ADTs in functional
  languages is usually justified by first representing ADTs by their
  Church encodings, and then reasoning type-independently about these
  encodings. This is typically justified by constructing a parametric
  model --- i.e, a model in which polymorphic functions preserve
  relations {\em \'a la} Reynolds --- for a suitable fragment of
  System F, demonstrating that an initial algebra exists for the
  positive type constructor corresponding to the functor underlying an
  ADT of interest, and showing that each such initial algebra is
  suitably isomorphic to its corresponding Church encoding. In fact,
  this isomorphism of initial algebras and their Church encodings is
  one of the ``litmus tests'' for the goodness of a parametric model.}

This approach works well for ADTs, which are always fixpoints of {\em
  first-order} functors, and whose Church encodings, which involve
quantification over only type variables, are always expressible in
System F. For example, $\mathtt{List\,A}$ is the fixpoint of the
first-order functor $F\,X = 1 + A \times X$ and has Church encoding
$\forall \alpha. \, \alpha \to (\mathtt{A} \to \alpha \to \alpha) \to
\alpha$. But despite Cardelli's~\cite{car97} claim that ``virtually
any basic type of interest can be encoded within F$_2$'' --- i.e.,
within System F --- non-ADT nested types cannot.  Not even our
prototypical nested type of perfect trees has a Church encoding
expressible in System F!  Indeed, $\mathtt{PTree\,A}$ cannot be
represented as the fixpoint of any {\em first-order} functor. However,
it can be seen as the instance at index $\mathtt{A}$ of the fixpoint
of the {\em higher-order} functor $H\,F\,A\,=\, (A \to F\,A) \to (F
\,(A \times A) \to F\,A) \to F\,A$. It thus has Church encoding
$\forall f.\, (\forall \alpha.\,\alpha \to f\alpha) \to (\forall
\alpha. \,f (\alpha \times \alpha) \to f\alpha) \to \forall \alpha.\,
f\alpha$, which requires quantification at the higher kind $* \to *$
for $f$. A similar situation obtains for any (non-ADT) nested
type. Unfortunately, higher-kinded quantification is not available in
System F, so if we want to reason type-independently about nested
types in a language based on it we have only two options: {\em
  i})\,move to an extension of System F, such as the higher-kinded
calculus F$_\omega$ or a dependent type theory, and reason via their
Church encodings in a known parametric model for that extension, or
{\em ii})\, add nested types to System F as primitives --- i.e., as
primitive type-level fixpoints --- and construct a parametric model
for the result.

Since the type systems of F$_\omega$ and dependent type theories are
designed to extend System F with far more than non-ADT data types, it
seems like serious overkill to pass to their parametric models to
reason about nested types in System F. Indeed, such calculi support
fundamentally new features that add complexity to their models that is
entirely unnecessary for reasoning about nested types. This paper
therefore pursues the second option above.  We first design a
Hindley-Milner-style calculus supporting primitive nested types,
together with primitive types of natural transformations representing
morphisms between them. Our calculus can express all nested types
appearing in the literature, including truly nested types.  At the
term-level, it supports primitive pattern matching, map functions, and
fold combinators for nested types.\footnote{We leave incorporating
  general term-level recursion to future work because, as
  Pitts~\cite{pit00} reminds us, ``it is hard to construct models of
  both impredicative polymorphism and fixpoint recursion''. In fact,
  as the development in this paper shows, constructing a parametric
  model even for our predicative calculus with primitive nested types
  --- and even without term-level fixpoints --- is already rather
  involved.  On the other hand, our calculus is strongly normalizing,
  so it perhaps edges us toward the kind of provably total practical
  programming language proposed in~\cite{wad89}.}  Our main
contribution is the construction of a parametric model for our
calculus. This is both delicate and challenging. To ensure the
existence of semantic fixpoints interpreting nested types, and thus to
establish a suitable Identity Extension Lemma, our type system must
explicitly track functoriality of types, and cocontinuity conditions
on the functors interpreting them must be appropriately threaded
throughout the model construction. Our model validates all standard
consequences of parametricity in the presence of primitive nested
types, including the isomorphism of primitive ADTs and their Church
encodings, and correctness of short cut fusion for nested types. The
relationship between naturality and parametricity has long been of
interest, and our inclusion of a primitive type of natural
transformations allows us to clearly delineate those consequences of
parametricity that follow from naturality, from those, such as short
cut fusion for nested types, that require the full power of
parametricity.
\end{comment}

\subsection{GADT Syntax}\label{sec:gadts}

{\color{blue} polynomial} GADTs relax the restriction on the type
instances appearing in a data type definition. The special form of
GADTs known as {\em nested types} allow the data constructors of a
GADT to take as arguments data whose types involve type instances of
the GADT other than the one being defined. For example, the 
definition 
\[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
introduces the data constructor(s $\mathsf{PLeaf :: a \to PTree\,a}$
and) $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
not only the invariant that all of the data in a structure of type
$\mathsf{PTree\,a}$ is of the same type $\mathsf{a}$, but also the
invariant that all perfect trees have lengths that are powers of
2. More general GADTs allow their constructors both to take as
arguments {\em and return as results} data whose types involve type
instances of the GADT other than the one being defined. An example is
the following GADT, given in Haskell syntax:
\[\begin{array}{l}
\mathsf{data\, Seq :: * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}\]
Since the return type of the data constructor $\mathsf{Pair}$ is not
simply a variable, $\mathsf{Seq}$ is not a nested type.

@@

\subsubsection{Church Encodings of GADTs}\label{sec:CEs}

@@

\subsubsection{GADTs as Primitive Data Types}\label{sec:prim-gadts}

@@

As we have seen in the previous Section~\ref{sec:reps}, GADTs
represented as Church encodings do not necessarily have a map or a
functorial semantics.  In particular, this means that we cannot
possibly derive naturality results for those types, event though their
semantics is parametric.  Indeed, even though naturality is sometimes
seen as a consequence of parametricity, that is only true in case we
have a functorial semantics and types supporting a map.  Still, GADTs
represented as Church encodings can still exhibit consequences of
paramentricity other than naturality, as they do admit parametric
semantics~\cite{atk12}.

>>>>

\section{Representations of Algebraic Data Types}\label{sec:adts}

@@

\subsection{Church Encodings of ADTs}

@@

\subsection{ADTs as Primitive Data Types}

@@

Primitive representations capture in syntax the fact that ADTs can be
considered as fixpoints. For example $\mathsf{List\,a}$ is indeed a
fixpoint of $\mathsf{F_{List\,a}}$, where $\mathsf{F_{List\,a}\,X} =
\mathsf{1 + a \times X}$ as determined by~\eqref{eq:list}, i.e., is
indeed such that
\[\mathsf{List\,a} = \mathsf{1 + a \times List\,a}\] 
since every element of $\mathsf{List\,a}$ is either empty or is
obtained by $\mathsf{Cons}$ing an element of type $\mathsf{a}$ onto an
already-existing structure of type $\mathsf{List\,a}$. In fact, this
fixpoint equation is just a rewriting of the Haskell data type
declaration for $\mathsf{List\,a}$. We therefore have that
$\mathsf{List\, a} = \mathsf{\mu F_{List\,a}}$ is modelled by $\mu
F_{\mathit{List}\,a}$, where the functor $F_{\mathit{List}\,a}$ models
the type constructor $\mathsf{F_{List\,a}}$. Similarly, the ADT
$\mathsf{Tree\,a}$ can be seen to be a fixpoint of
$\mathsf{F_{Tree\,a}}$, where $\mathsf{F_{Tree\,a}\,X} = \mathsf{a + X
  \times a \times X}$ as determined by~\eqref{eq:tree}, so that
$\mathsf{Tree\, a} = \mathsf{ \mu F_{Tree\,a}}$ is modelled by $\mu
F_{\mathit{Tree}\,a}$, where $F_{\mathit{Tree}\,a} $ models
$\mathsf{F_{Tree\,a}}$.

These kinds of fixpoint equations are entirely sensible at the level
of types. But to ensure that the syntactic fixpoint representing an
ADT actually denotes a semantic object computed as a real-live
fixpoint, the semantic fixpoint calculation must converge.  If we
interpret our types as sets, then this means that the fixpoint being
taken must be of a {\em functor} on the category $\mathit{Set}$ of
sets and functions between them, rather than of a mere function
between sets~\cite{tfca}.  That is, the function $F$ interpreting the
type constructor $\mathsf{F}$ constructing the body of a syntactic
fixpoint must not only have an action on sets, but must also have a
{\em functorial action} on functions between sets. Reflecting this
requirement back into syntax gives that $\mathsf{F}$ must support a
suitable\footnote{The $\mathsf{map}$ function for $\mathsf{F}$ should
  satisfy syntactic reflections of the functoriality properties ---
  i.e., should preserve identities and compositions --- in
  $\mathit{Set}$ even though there is no mechanism in Haskell for
  enforcing this.} $\mathsf{map}$ function, i.e., must be an instance
of Haskell's $\mathsf{Functor}$ class.

Requiring $F$ to be a functor will ensure that the interpretation $\mu
F$ of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure
that $\mu F$ is itself a functor, so that the type constructor
$\mathsf{T}$ associated with the ADT $\mathsf{T\,a}$ supports a
$\mathsf{map}$ function as well, $F$ must again be the fixpoint $\mu
H$ of a functor $H$. In this case, however, $H$ must be a functor on
the category $\mathit{Func(Set)}$ of functors and natural
transformations on $\mathit{Set}$, i.e., $H$ must be a {\em
  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu F)\,a} =
\mathsf{(\mu\,(\mu H))\,a}$ for the ``type constructor constructor''
$\mathsf{H}$ that supports suitable\footnote{{\color{blue} The
    $\mathsf{map}$ function for $\mathsf{H}$ should satisfy syntactic
    reflections of the functoriality properties in
    $\mathit{Func(Set)}$ --- i.e., should preserve identity natural
    transformations and composition of natural transformations --- and
    the $\mathsf{map}$ function for $\mathsf{H\,F}$ should satisfy
    syntactic reflections of the functoriality properties in
    $\mathit{Set}$ as in the previous footnote. Again, there is no
    mechanism in Haskell for enforcing this.}} $\mathsf{map}$
functions.  {\color{blue} Mention naturality?}

%class HFunctor f where
%  ffmap :: Functor g => (a -> b) -> f g a -> f g b
%  hfmap :: (g :~> h) -> f g :~> f h
%
%  type (~>) f g = forall x. f x -> g x

A concrete example is given by the ADT $\mathsf{List\,a}$. This type
is is modelled as the fixpoint $\mu F_{\mathit{List\,a}}$ of the
first-order functor whose action on sets is given by
$F_{\mathit{List}\,a}\,X = 1 + a \times X$ and whose action on
functions is given by $F_{\mathit{List}\,a}\,f = 1 + \id_a \times
f$. The type constructor $\mathsf{List}$ is modelled by the
higher-order functor $H$ whose action on a functor $F$ is given by the
functor $H\,F$ whose actions on sets and functions between them are
given by $H\,F\,X = 1 + X \times F\,X$ and $H\,F\,f = \id_1 + f \times
F\,f$, respectively, and whose action on a natural transformation
$\eta$ is the natural transformation given by $H\,\eta = \id_1 + \eta
\times F\,\eta$. The type constructor $\mathsf{List}$ is then
modelled as the functor $\mu H$. Reflecting the functorial action of
$\mu H$ back into syntax gives Haskell's built-in $\mathsf{map}$
function as the type-and-data-uniform, structure-preserving,
data-changing function associated with $\mathsf{List}$. {\color{blue}
  HOW?}

In a parametric model, short cut fusion can be used to show that the
Church encoding and the primitive representation of an ADT are
isomorphic as sets. Note, however, that whereas it is impossible even
to {\em state} the rearrange-transform property for Church encodings
of ADTs, it is simply the reflection back into syntax of the
{\color{blue} naturality property - right?} of the functorial action
of the interpretation of the type constructor associated with an ADT's
primitive representation.

\section{Representations of GADTs}\label{sec:gadts}

{\color{red} HERE!}

GADTs relax the restriction on the type instances appearing in a data
type definition. The special form of GADTs known as {\em nested types}
allow the data constructors of a GADT to take as arguments data whose
types involve type instances of the GADT other than the one being
defined. For example, the definition
\[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
introduces the data constructor(s $\mathsf{PLeaf :: a \to PTree\,a}$
and) $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
not only the invariant that all of the data in a structure of type
$\mathsf{PTree\,a}$ is of the same type $\mathsf{a}$, but also the
invariant that all perfect trees have lengths that are powers of
2. More general GADTs allow their constructors both to take as
arguments {\em and return as results} data whose types involve type
instances of the GADT other than the one being defined. An example is
the following GADT, given in Haskell syntax:
\[\begin{array}{l}
\mathsf{data\, Seq :: * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}\]
Since the return type of the data constructor $\mathsf{Pair}$ is not
simply a variable, $\mathsf{Seq}$ is not a nested type.

@@

\subsection{Church Encodings of GADTs}\label{sec:CEs}

@@

\subsection{GADTs as Primitive Data Types}\label{sec:prim-gadts}

@@

As we have seen in the previous Section~\ref{sec:adts}, GADTs
represented as Church encodings do not necessarily have a map or a
functorial semantics.  In particular, this means that we cannot
possibly derive naturality results for those types, event though their
semantics is parametric.  Indeed, even though naturality is sometimes
seen as a consequence of parametricity, that is only true in case we
have a functorial semantics and types supporting a map.  Still, GADTs
represented as Church encodings can still exhibit consequences of
paramentricity other than naturality, as they do admit parametric
semantics~\cite{atk12}.
